Modeling HEPES buffer with a charge-balance approach

### 2023-11-18 manuscript.Rmd
Least-squares solution of charge-balance equations
Instead of solving the charge-balance equation at every known SID, we
can use the least squares method to solve them all together, minimising
the sum of squared change imbalances.

This gives us worse results due to a different error structure of the
model, but may be worth noting.

\begin{equation}
[\mathrm{H}^{+}] + \mathrm{SID} - \frac{K_w}{[\mathrm{H}^{+}]}
+ \mathrm{I} \cdot 2 + \mathrm{II} - \mathrm{IV} = 0
\end{equation}

@@

```{r}
pH.wrapper <- function(SID, HEPTOT, pK1, pK2, pK3)
 vapply(SID, function(SID_i)
  -log10(uniroot(HEPESFUNC, c(1e-20,1), tol = 1e-20, maxiter = 1e4,
                 HEPTOT = HEPTOT, SID = SID_i,
                 pK1 = pK1, pK2 = pK2, pK3 = pK3)$root),
  numeric(1)
 )
```

@@

>>>>

\begin{equation}
[\mathrm{H}^{+}] + \mathrm{SID} - \frac{K_w}{[\mathrm{H}^{+}]}
+ \mathrm{I} \cdot 2 + \mathrm{II} - \mathrm{IV} = 0
(\#eq:charge-balance)
\end{equation}

@@

```{r}
pH.wrapper <- function(SID, HEPTOT, pK1, pK2, pK3)
 mapply(function(SID, HEPTOT)
  -log10(uniroot(HEPESFUNC, c(1e-20,1), tol = 1e-20, maxiter = 1e4,
                 HEPTOT = HEPTOT, SID = SID,
                 pK1 = pK1, pK2 = pK2, pK3 = pK3)$root),
  SID = SID, HEPTOT = HEPTOT
 )
```

@@

### Solving the charge-balance equations in the least squares sense

Yet another way to model the data is to plug the experimental values of
$\mathrm{SID}$ and measured $\mathrm{pH}$ straight into the
charge-balance equation \@ref(eq:charge-balance), without going through
the intermediate step of finding the negative logaritm of the root of
the function. This changes the problem: instead of
$\min_{\mathrm{p}K_{a,1}, \mathrm{p}K_{a,2}, \mathrm{p}K_{a,3}}
\sum_i(\mathrm p\hat{\mathrm H}(\mathrm{SID}_i, \mathrm{p}K_{a,1},
\mathrm{p}K_{a,2}, \mathrm{p}K_{a,3}) - \mathrm{pH}_i)^2$, we are now
minimising the sum of squared residuals from eq.
\@ref(eq:charge-balance) itself:

```{r}
load('HEPES.RData') # provides the DD data.frame
charge_balance_fn <- function() NULL
formals(charge_balance_fn) <- formals(HEPESFUNC)
body(charge_balance_fn) <- local({
	XX <- quote(H^3/(10^-pK3*10^-pK2*10^-pK1)+H^2/(10^-pK3*10^-pK2)+H/(10^-pK3))
	IV <- substitute(HEPTOT/(1+XX))
	I <- substitute(IV*H^3/(10^-pK3*10^-pK2*10^-pK1))
	II <- substitute(IV*H^2/(10^-pK3*10^-pK2))
	III <- substitute(IV*H/10^-pK3)
	nlsr::codeDeriv(
		substitute(H - kw/H + SID + I*2 + II - IV),
		c('pK1', 'pK2', 'pK3')
	)
})

loss <- function(x) charge_balance_fn(
	H = 10^-DD$OBS, SID = DD$SID, HEPTOT = DD$HEPES,
	pK1 = x['pK1'], pK2 = x['pK2'], pK3 = x['pK3']
)
(m_cb <- nlsr::nlfb(
	start = c(pK1 = -1, pK2 = 3, pK3 = 7.55),
	resfn = loss, jacfn = loss,
	lower = -3, upper = 9
))
```

While this avoids finding the roots of the charge-balance equation, the
implied residual structure (normally-distributed errors in the
$\mathrm{pH}$ measurements in the original model vs.
normally-distributed charge imbalances in concentration units in the
current model) is different, and so are the resulting estimates:

```{r}
(m_rf <- nlsr::nlxb(
	OBS ~ pH.wrapper(SID, HEPES, pK1, pK2, pK3),
	data = DD, start = c(pK1 = -1, pK2 = 3, pK3 = 7.55),
	lower = -3, upper = 9,
	control = list(japprox = 'jacentral')
))
```
